---
title: "Doggone Data (Midterm Project)"
author: "Nicholas Johnson & Samuel Rasely"
date: "2/16/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

As a way of showcasing what we have learned so far, we were tasked with cleaning and exploring one of the data sets provided to us. In our case, we selected the data set pertaining to dogs that are available for adoption in the United States.

We will be utilizing RStudio to perform cleaning and exploratory analysis on the data set. And, in order to make our work easier to understand, the steps we took will be broken down into the following:

1. Required Packages
2. Data Preparation and Cleaning
3. Proposed Exploratory Analysis

## Required Packages
``` {r message = FALSE}
library("tidyverse")# To help tidy up the data
library("readr")    # To import .csv files in a more feature rich way
library("here")     # To make it easier to work collaboratively on the project
library("dplyr")    # For data/dataframe manipulation
```

## Data Preparation and Cleaning

As the data was provided to us, the source is [on our class' Canvas page](https://canvas.ewu.edu/courses/1540406/files/87301233?module_item_id=14963744)

The data is broken down into three .csv files:

* dog_adoptable.csv
* dog_descriptions.csv
* dog_destination.csv

The data was collected on 9/20/2019 from various dog adoption organizations across the US.

### First, setting our work directory and retrieving the data
```{r, warning = FALSE, message = FALSE, results = FALSE}
here::i_am("Final.Rproj")
here() # set current directory to top-level of project 

# load dog_adoptable, dog_descriptions and dog_destination
dog_adoptable <- read_csv("data/raw/dog_adoptable.csv")
dog_descriptions <- read_csv("data/raw/dog_descriptions.csv")
dog_destination <- read_csv("data/raw/dog_destination.csv")

# loading the variable information for each table
dog_adoptable_variable_table <- read_csv("data/raw/dog_adoptable_variable_table.csv")
dog_descriptions_variable_table <- read_csv("data/raw/dog_descriptions_variable_table.csv")
dog_destination_variable_table <- read_csv("data/raw/dog_destination_variable_table.csv")
```

<br>

#### Now we can see the variable descriptions for the data we've loaded.

<br>

##### dog_adoptable variables and descriptions
```{r, echo = FALSE}
knitr::kable(dog_adoptable_variable_table, col.names = gsub("[.]", " ", names(dog_adoptable_variable_table)))
```

<br>

##### dog_adoptable variables and descriptions
```{r, echo = FALSE}
knitr::kable(dog_descriptions_variable_table, col.names = gsub("[.]", " ", names(dog_adoptable_variable_table)))
```

<br>

##### dog_adoptable variables and descriptions
```{r, echo = FALSE}
knitr::kable(dog_destination_variable_table, col.names = gsub("[.]", " ", names(dog_adoptable_variable_table)))
```

<br>

#### Prepping to clean
Now, after reviewing the data set's variables and their meanings we had to decide how to clean them. We found that rather than going through and cleaning ever single variable, it worked better for us to clean the data while keeping in mind what questions we wanted to be able to answer using the data.

<br>

Some of the questions we wanted to answer were:

* How many adoptable dogs were there in each state?
* What is the most common breed of the adoptable dogs in each state?
* How long has each dog been up for adoption?

<br>

Figuring out what questions we wanted to ask first helped us shave down the number of variables we needed to keep (and clean) in the data set.

<br>

#### Cleaning the data

##### dog_adoptable
```{r, warning = FALSE, message = FALSE, results = FALSE}
# dog_adoptable

# update field inUS to snake_case in_us
dog_adoptable <- rename(dog_adoptable, in_us = inUS)

# filter to just true for in_us
dog_adoptable <- filter(dog_adoptable, in_us == TRUE)

# drop field in_us as all records have same value
dog_adoptable <- select(dog_adoptable, !in_us)

# replace all NA with 0
dog_adoptable <- mutate_all(dog_adoptable, ~replace(., is.na(.), 0))

# rename location to state
dog_adoptable <- rename(dog_adoptable, state = location)
```

<br>

|      Starting with the dog_adoptable data, we renamed the 'inUS' variable to in_us, however, since all the observations are located in the US we ended up discarding the 'in_us' column. We also renamed the 'location' variable to 'state' as we felt that title more accurately reflected the values in the column. And, while there are outliers in 'exported', 'imported' and 'total', we found them all of them to be believable values.

<br>

##### dog_destination
```{r, warning = FALSE, message = FALSE, results = FALSE}
# dog_destination (removed)
```

<br>

|       The dog_destination data was by and far the worst one. It contained a minimal amount of additional information, and the information it did contain was contradictory. For example, the 'contact_city' and 'contact_state' would contradict where the 'found' variable, and the variables 'remove' and 'still_there' did not reflect the dog's change of location.

Not only this, but the values in the 'found' column were inconsistent. It lists countries, counties, cities, and nonsensical values such as 'Sunday 10am' or 'Glaucoma'. As a result, we decided to drop the table completely.

<br>

##### dog_descriptions
```{r, warning = FALSE, message = FALSE, results = FALSE}
# dog_descriptions

# drop stateQ as is only "The state abbreviation queried in the API to return this result " 
dog_descriptions <- select(dog_descriptions, !stateQ)

# drop status field as all are dogs adoptable
dog_descriptions <- select(dog_descriptions, !status)

# drop species field as all are dogs so adds no value.
dog_descriptions <- select(dog_descriptions, !species)

# drop type field as all are dogs so adds no value.3 are NA but confirmed they are dogs with their description
dog_descriptions <- select(dog_descriptions, !type)

# drop photo as all are NA
dog_descriptions <- select(dog_descriptions, !photo)

# drop name as useless
dog_descriptions <- select(dog_descriptions, !name)

# drop tags as useless
dog_descriptions <- select(dog_descriptions, !tags)

# drop description as useless
dog_descriptions <- select(dog_descriptions, !description)

# drop declawed as all are NA
dog_descriptions <- select(dog_descriptions, !declawed)

# drop contact_country as all are in the US, some have state or zip here by error
dog_descriptions <- filter(dog_descriptions, contact_country == "US")
dog_descriptions <- select(dog_descriptions, !contact_country)
```

<br>

|       Here we went through dog_descriptions and dropped columns that either contained information that we did not need, or had no useful values in them. For example, all of the values for 'declawed' were 'NA' so we dropped the column.

<br>

##### dog_destination (continued)
```{r, warning = FALSE, message = FALSE, results = FALSE}
dog_descriptions <- mutate(dog_descriptions, posted_date = as.Date(posted))
dog_descriptions <- mutate(dog_descriptions, accessed_date = as.Date(accessed, "%d/%m/%Y"))
dog_descriptions <- mutate(dog_descriptions, days_in_shelter = 
                             as.numeric(difftime(dog_descriptions$accessed_date, 
                                                 dog_descriptions$posted_date , units = c("days"))))
dog_descriptions <- select(dog_descriptions, !c(posted, posted_date, accessed, accessed_date))

# fix zip na, all are in Boston 02108
dog_descriptions <- mutate_at(dog_descriptions, vars("contact_zip"), ~replace(., is.na(.), 02108))

# TODO: why is this not padding zip with 0s?
dog_descriptions <- mutate(dog_descriptions,
          zip = str_pad(string = contact_zip,
                              width = 5,
                              side = "left",
                              pad = "0"))

#rename to city state and zip
dog_descriptions <- select(dog_descriptions, !contact_zip)
dog_descriptions <- rename(dog_descriptions, city = contact_city)
dog_descriptions <- rename(dog_descriptions, state = contact_state)

# state abbreviation to full names
state.abb.and.name <- tibble(state.abb, state.name)

# left join for state information
dog_descriptions <- left_join(dog_descriptions, state.abb.and.name, by = c("state" = "state.abb"))
```

<br>

In this section we renamed, and mutated some variables. 

<br>

##### dog_destination (continued)
```{r, warning = FALSE, message = FALSE, results = FALSE}
# breed_secondary
dog_descriptions$breed_secondary[is.na(dog_descriptions$breed_secondary)] <- "NONE / UNKNOWN"

# color_primary
dog_descriptions$color_primary[is.na(dog_descriptions$color_primary)] <- "OTHER"

# color_secondary
dog_descriptions$color_secondary[is.na(dog_descriptions$color_secondary)] <- "NONE / OTHER"

# color_tertiary
dog_descriptions$color_tertiary[is.na(dog_descriptions$color_tertiary)] <- "NONE / OTHER"

# coat
dog_descriptions$coat[is.na(dog_descriptions$coat)] <- "OTHER"
```

<br>

|       Here we took care of the remaining 'NA' values in the character type columns. As opposed to having colors or breeds being listed as 'NA' we decided that a value of 'NONE / UNKOWN', 'NONE / OTHER', or 'OTHER' were more appropriate. 
|       The reasoning behind our logic was that a dog could be purebred or a mix of different breeds with no distinguishable second primary breed. We applied the same logic to the dog coloration and coat since sometimes there is no obviously dominant coloration or appropriate descriptor for what the dog might be.

<br>

#### Below are small chunks of the cleaned data

<br>

##### dog_adoptable after cleaning
```{r, warning = FALSE, message = FALSE, echo = FALSE}

knitr::kable(head(dog_adoptable), col.names = gsub("[.]", " ", names(dog_adoptable)))
```

<br>

##### dog_descriptions after cleaning
```{r, warning = FALSE, message = FALSE, echo = FALSE}

knitr::kable(head(dog_descriptions), col.names = gsub("[.]", " ", names(dog_descriptions)))
```
<br>

## Proposed Exploratory Analysis
|       We mentioned some of the questions we wanted to ask earlier and our proposed analysis is to follow through and answer those questions. Towards that end, we would like to visualize which states have the most dogs available for adoption, and break down those states or regions into the most common breeds of dogs that can be adopted.

|       Following up on that, we would also like to see if there is a connection in specific traits that might contribute towards the length of time a dog spends being available for adoption (such as being good around kids, cats, or other dogs).








